Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling

### 2008-02-02 paper.tex
Added a bad-ass footnote about flow pairing.
Working on adding the flow pairing stuff into the representation section.

The traffic workload pattern in a network is simply a collection of packets between hosts in the network sent at certain times. It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers. Such a sequence of packets is called a \textit{flow}. The traffic pattern is characterized completely by the detailed behavior of it's flows. The behavior of a flow is defined by six properties: its source and destination nodes, its start and end times, and the specific sizes of and intervals between its constituent packets. We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior. The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, the rows of which are the behavior vectors of the flows.

@@

The {packets-bytes-duration} model representation is clearly overly simplistic and very incomplete. (Yet, it is more detailed than most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow. Accordingly, we represent a flow's behavior by these precise characteristics. If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\setx{\vec{d}_i}\:$\=$=\setx{d_{ij}}\:$\=\kill
\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$=\setx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$=\setx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$=\setx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$=\setx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}

@@

Let $n$ be the number of nodes in the given traffic collection. We map nodes onto the standard basis vectors of $\R^n$: if the $j$th node is the source of the the $i$th flow, then $\vec{s}_i=\mathbf{e}_j$ (the vector with 1 at index $j$ and zeros elsewhere); if the $k$th node is the destination, then $\vec{d}_i=\mathbf{e}_k$. This may seem like a rather expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$. The advantage becomes clear when we consider vector operations. The sum of the source indices over a collection is meaningless; the sum of the the source vectors, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.

@@

We also represent start and end times using indicator vectors. Time, however, is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$. We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
\begin{align}
\Qt(t) &= \max\set{1,\ceil{\fracx{d_t t}{t_{\max}}}}.
\end{align}
Let $j$ and $k$ be the quantized start and end indices of a flow. Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$. Again, this may seem excessively complex, but the advantage becomes clear when take aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.

@@

The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$, whose $i$th element, $z_i$, is the number of packets in the flow with $i$ bytes of payload data. The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes. The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we much map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$. In our experiments, we use the following adaptive quantization function:
\begin{align}
\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^{\frac{1}{2}}}}.
\end{align}
This specific function was chosen because, given the values of $d_v$, $v_{\min}$, and $v_{\max}$ in our experiments, it provides millisecond-scale to second-scale resolution for intervals on the orders of $v_{\min}$ to $v_{\max}$, respectively, adapting resolution smoothly between these extremes.
%The values we use are $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and using $d_v=1500$ to match the dimensionally of the packet size representation.
The summation of behavior vectors over multiple flows yields a joint histogram giving the pooled frequencies of all packet sizes and inter-packet intervals.

\section{Derivative Representations of Network Behavior}

Now that we have defined our representation of network behavior, we derive useful expressions for commonly used network properties in terms of these basic representations.

@@

The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t\in\R^{n \times n}$. The entry at row $i$ and column $j$ in this matrix gives the number of flows from

>>>>

The traffic workload pattern in a network is simply a collection of packets between hosts in the network sent at certain times. It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers. Such a sequence of packets is called a \textit{flow}. The traffic pattern is characterized completely by the detailed behavior of it's flows. The behavior of a flow is defined by seven properties: its source and destination nodes; its start and end times; the specific sizes of and intervals between its constituent packets; and finally, whether it is a response, paired with another flow.\footnote{Flow pairing is typically not addressed in traffic models. However, the most common transport protocol, {\scriptsize{TCP}}, is full-duplex, exhibiting tightly coupled pairs of client-server flows. Therefore, accurate representation of traffic behavior requires knowing which flows are request-response pairs. Moreover, although they are not strictly full-duplex, it is generally appropriate to view {\scriptsize{UDP}} and even {\scriptsize{ICMP}} flows as paired, since they are typically paired. For example, {\scriptsize{DNS}} responses are naturally associated with the corresponding request, and will have symmetrical {\scriptsize{IP}} addresses and port numbers.} We propose to represent the behavior of each flow in terms of seven vectors---one for each aspect of its behavior. The aggregate behavior of all the flows in a workload pattern may then be represented by these matrices, the rows of which are the behavior vectors of the flows.

@@

The {packets-bytes-duration} model representation is clearly overly simplistic and very incomplete. (Yet, it is more detailed than most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, empirical distributions of packet sizes and inter-packet intervals for each flow, and which other flow (if any) it is paired with. If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with seven matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\setx{\vec{d}_i}\:$\=$=\setx{d_{ij}}\:$\=\kill
\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$=\setx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> flow pairs: \>$P$\>$=\setx{\vec{p}_i}$\>$=\setx{p_{ij}}$\>$\in\R^{f \times f}$,\\
\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$=\setx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$=\setx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$=\setx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}

@@

Let $n$ be the number of nodes in the given traffic collection. We map nodes onto the standard basis vectors of $\R^n$: if the $j$th node is the source of the the $i$th flow, then $\vec{s}_i=\mathbf{e}_j$ (the vector with 1 at index $j$ and zeros elsewhere); if the $k$th node is the destination, then $\vec{d}_i=\mathbf{e}_k$. This may seem like an overly expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$. The advantage becomes clear when we consider vector operations. The sum of the source indices over a collection is meaningless; the sum of the the source vectors, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.

\subsection{Flow Pairing: Clients \& Servers}

@@

We represent start and end times using indicator vectors. Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$. We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
\begin{align}
\Qt(t) &= \max\set{1,\ceil{\fracx{d_t t}{t_{\max}}}}.
\end{align}
Let $j$ and $k$ be the quantized start and end indices of a flow. Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$. Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.

@@

The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$. Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data. The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes. The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we much map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$. In our experiments, we use the following adaptive quantization function:
\begin{align}
\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^{\frac{1}{2}}}}.
\end{align}
This specific function was chosen because, given the values of $d_v$, $v_{\min}$, and $v_{\max}$ in our experiments, it provides millisecond-scale to second-scale resolution for intervals on the orders of $v_{\min}$ to $v_{\max}$, respectively, adapting resolution smoothly between these extremes.
%The values we use are $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and using $d_v=1500$ to match the dimensionally of the packet size representation.
The summation of behavior vectors over multiple flows yields a joint histogram giving the pooled frequencies of all packet sizes and inter-packet intervals.

\section{Expressions for Common Workload Properties}

Having defined our representation of network behavior, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.

@@

The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t$: $(SD^t)_{ij}$ is the number flows from node $i$ to node $j$.
